---
title: "Function Scoping in Depth"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

Using the makeVector() Example:  
  
```{r}
makeVector <- function(x = numeric()) {
    m <- NULL
    set <- function(y) {
            x <<- y
            m <<- NULL
    }
    get <- function() x
    setmean <- function(mean) m <<- mean
    getmean <- function() m
    list(set = set, get = get,
         setmean = setmean,
         getmean = getmean)
}
#
#
cachemean <- function(x, ...) {
    m <- x$getmean()
    if(!is.null(m)) {
            message("getting cached data")
            return(m)
    }
    data <- x$get()
    m <- mean(data, ...)
    x$setmean(m)
    m
}
```

`makeVector()` returns a set of functions in a list:  

- `set()`
- `get()`
- `setmean()`
- `getmean()`
  
Important points:  

- Each function has its own environment in R.  
- When calling `myVector <- makeVector()`, `myVector` has access to the entire `makeVector` environment.  
- `myVector` contains pointers to functions that are within the `makeVector()` environment which prevents the memory consumed by `makeVector()` from being discarded.  

Upon initialization, `myVector()` also adds two data objects to the environment:  

- `x` 
  - The argument initialized on the original function call (initialized as a function argument)
    - By default it is initialized as `numeric()` so that `data <- x$get()` doesn't result in an error from a missing argument.
  - It is accessible by subsequent calls to functions on `myVector` such as `myVector$get()`
  - This code works without having to issue `myVector$set()`
  
- `m`
  - Set to NULL
  
Mutator and Accessor methods:  

- Typical behaviours for data elements within object-oriented programs.
- Also called "getters and setters".
  - "Getters" retrieve (access) data within an object
  - "Setters" set (mutate) the data values within an object
  
```{r}
set <- function(y) {
        x <<- y
        m <<- NULL
}
```

- `set()` takes an argument y (It is labelled differently from `x` to make the code easier to understand).
- `<<-` denotes that the object on the left side of the operator (`x`) is in the parent environment.  

`set()`:  

1. Assigns the input argument to the `x` object in the parent environment (makeVector environment)
2. Assigns the value of NULL to the `m` object in the parent environment.
    - This clears any value of `m` previously cached by `cachemean()`, forcing subsequent calls to `cachemean()` to recalculate the mean rather than retrieving the wrong value from the cache.
    
`set()` performs the same operations as the first two lines of the main function.
  
`set()` is called outside of both functions to reset value with a new vector.

```{r}
get <- function() x
```

- since `x` is not defined within `get()`, R retrieves it from the parent environment of `makeVector()`
- `function() x` == `function() {x}`
- Calling this function with no argument just returns `x`

```{r}
setmean <- function(mean) m <<- mean
```

- `setmean()` uses the `<<-` operator to assign the input argument to the value of `m` in the parent environment.
  - This is required because `m` is defined in the parent environment and we need to access it after `setmean()` completes

```{r}
getmean <- function() m
```

- This operates in much the same way as `get()` and for the same purpose.
  
Now we have getters and setters defined for both of the data objects within the makeVector() object.

```{r}
    list(set = set,          # gives the name 'set' to the set() function defined above
         get = get,          # gives the name 'get' to the get() function defined above
         setmean = setmean,  # gives the name 'setmean' to the setmean() function defined above
         getmean = getmean)  # gives the name 'getmean' to the getmean() function defined above
```

- Here we assign each of the functions as an element within a `list()` and return it to the parent environment.
- This list is a fully formed object of type `makeVector()`
- Each element is this list is also named:
  - `elementName = value` syntax
  - This allows us to use the `$` form of the extract operator to access the functions by name rather than using the `[[` form of the extract operator (i.e. `myVector[[2]]()`) to get the contents of the vector.

`cachemean()`:

- Requires an input argument of type `makeVector()` (a regular vector will be unable to access `$getmean()` because `$` does not work with atomic vectors).
- `cachemean()` is the actual function that populates/retrieves with the tools provided by `makeVector()`

```{r}
cachemean <- function(x, ...) {
  ...
}
```

- The function takes in a single argument `x` as well as possible other arguments
- The input of `cachemean()` is the output of `makeVector()`

```{r}
m <- x$getmean()
```

- This attempts to retrive a mean from the object passed in as the argument.

```{r}
if(!is.null(m)) {
        message("getting cached data")
        return(m)
}
```

- If `!is.null(m)` is `TRUE`, `m` is returned from the vector.
- If `!is.null(m)` is `FALSE`:

```{r}
data <- x$get()
m <- mean(data, ...)
x$setmean(m)
m
```

- `cachemean()` gets the vector from the input object
- Calculates a `mean()`
- Uses `setmean()` on the input object to set the mean in the input object
- Returns the value of the mean to the parent environment by printing the mean object.

`cachemean()` is the only place where the `mean()` function is executed.