---
title: "Week 3 - Loop Functions and Debugging"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```
### Loop Functions
- `lapply`: Loop over a list and evaluate a function on each element
- `sapply`: Same as lapply but try to simplify the result
- `apply`: Apply a function over the margins of an array
- `tapply`: Apply a function over subsets of a vector (table)
- `mapply`: Multivariate version of `lapply`
  
#### lapply
- Takes in 3 arguments:
  1. a list `x`
  2. a function `FUN`
  3. other arguments via `...` 
- If `X` is not a list, it will be coerced to alist using as.list.
- Always returns a list, regardless of the class of the input.
- The actual looping is done in C code
  
```{r}
function (X, FUN, ...) {
        FUN <- match.fun(FUN)
        if(!is.vector(X) || is.object(X))
            X <- as.list(X)
        .Internal(lapply(X, FUN))
}
<bytecode: 0x7ff7a1951c00>
<environment: namespace:base>
```

```{r}
x <- list(a = 1:5, b = rnorm(10))
lapply(x, mean)
$a
[1] 3

$b
[1] 0.0296824
```

```{r}
x <- list(a = 1:4, b = rnorm(10), c = rnorm(20, 1), d = rnorm(100, 5))
lapply(x, mean)
$a
[1] 2.5

$b
[1] 0.06082667

$c
[1] 1.467083

$d
[1] 5.074749
```

```{r}
x <- 1:4
lapply(x, runif) #x amount of uniform distribution values, between 0 and 1
[[1]]
[1] 0.2675082

[[2]]
[1] 0.2186453 0.5167968

[[3]]
[1] 0.2689506 0.1811683 0.5185761

[[4]]
[1] 0.5627829 0.1291569 0.2563676 0.7179353
```


```{r}
x <- 1:4
lapply(x, runif, min = 0, max = 10) #x amount of uniform distribution values, between 0 and 10
[[1]]
[1] 7.108169

[[2]]
[1] 7.10346 4.74639

[[3]]
[1] 9.688282 5.138259 9.005831

[[4]]
[1] 5.9789127 9.6521355 2.5745983 0.8854886
```

```{r}
x <- list(a = matrix(1:4, 2, 2), b = matrix(1:6, 3, 2))
x
$a
     [,1] [,2]
[1,]    1    3
[2,]    2    4

$b
     [,1] [,2]
[1,]    1    4
[2,]    2    5
[3,]    3    6

lapply(x, function(elt) elt[,1]) 
$a
[1] 1 2

$b
[1] 1 2 3

#This function only exists in lapply to pull the first column of each matrix
#x becomes inserted for "elt"
```

#### sapply
- If the result is a list where every element is length 1, then a vector is returned.
- If the result is a list where every element is avector of the same length (>1), a matrix is returned.
- If it can't figure things out, a list is returned.  

Remember:

```{r}
x <- list(a = 1:4, b = rnorm(10), c = rnorm(20, 1), d = rnorm(100, 5))
lapply(x, mean)
$a
[1] 2.5

$b
[1] 0.06082667

$c
[1] 1.467083

$d
[1] 5.074749
```
  
Alternatively:

```{r}
sapply(x, mean)
        a         b         c         d 
2.5000000 0.2139836 0.4600913 4.9709709 

mean(x)
[1] NA
```
  
#### apply
- Most often used to apply a function to the rows or columns of a matrix
- Can be used with general arrays (e.g. taking the average of an array of matrices)
- Not really faster than writing a loop, but it works in a line!

```{r}
str(aapply(function (X, MARGIN, FUN, ...)
```
- `X` is an array
- `MARGIN` is an integer vector indicating which margins should be "retained".
- `FUN` is a function to be applied
- `...` is for other arguments to be passed to `FUN`

```{r}
x <- matrix(rnorm(200), 20, 10)
apply(x, 2 ,mean) #presrve all the columns(2) collapses the rows
 [1] -0.28626874  0.19856511 -0.31688305 -0.02731786  0.02768524
 [6] -0.02500611  0.18800750  0.05201392  0.37343831 -0.19715015

apply(x, 1, sum) #preserve all the rows(1), collapse all the columns
 [1] -4.7336987  0.4856736  3.0622113  3.1236871 -4.9419170 -1.5839072
 [7]  0.4928710 -5.1211304 -1.3053737 -5.1036233  0.9341416 -0.8665630
[13]  2.3178583 -0.6342967  5.5691680  0.5782443  2.9294297  2.2108695
[19] -0.4902686  2.8183079
```

```{r}
rowSums = apply(x, 1, sum)
rowMeans = apply(x, 1, mean)
colSums = apply(x, 2, sum)
colMeans = apply(x, 2, mean)
```

- These shortcuts are much faster

```{r}
x <- matrix(rnorm(200), 20, 10)
apply(x, 1, quantile, probs = c(0.25, 0.75))
 [,1]       [,2]        [,3]        [,4]       [,5]
25% -0.5809978 -0.8237116 -0.62387762 -0.68944995 -0.5281022
75%  0.5865756  0.3815251  0.08641744  0.09694716  0.2109390
          [,6]       [,7]       [,8]        [,9]      [,10]      [,11]
25% -0.8184700 -0.3546380 -0.2197593 -0.55889314 -1.0146341 -0.2654860
75%  0.5549174  0.3347432  1.0386384 -0.05015169  0.5391982  0.8761715
         [,12]      [,13]        [,14]      [,15]        [,16]
25% -0.6949905 -0.4205925 -0.009323959 -0.5887117 -0.968603750
75%  0.9399615  0.0426014  0.399888547  0.7000215 -0.009519585
         [,17]      [,18]       [,19]     [,20]
25% -0.5613768 -1.0133901 -1.25971679 -0.917481
75%  0.9533017  0.1325692  0.00429806  1.342292
```

```{r}
a <- array(rnorm(2 * 2 * 10), c(2, 2, 10))
apply(a, c(1, 2), mean)
          [,1]       [,2]
[1,] 0.1473825 -0.2412296
[2,] 0.4247021 -0.2485657

rowMeans(a, dims = 2) #collapse to 2-dimensions
          [,1]       [,2]
[1,] 0.1473825 -0.2412296
[2,] 0.4247021 -0.2485657
```


#### mapply
- A multivariate apply which applies a function in parallel over a set of arguments.
```{r}
str(mapply)
function (FUN, ..., MoreArgs = NULL, SIMPLIFY = TRUE, USE.NAMES = TRUE)
```
- `FUN` is a function to apply
- `...` contains arguments to apply over
- `MoreArgs` is a list of other arguments to `FUN`.
- `SIMPLIFY` indicates whether the results should be simplified

```{r}
list(rep(1, 4), rep(2, 3), rep(3, 2), rep(4, 1))

vs.

mapply(rep, 1:4, 4:1)
[[1]]
[1] 1 1 1 1

[[2]]
[1] 2 2 2

[[3]]
[1] 3 3

[[4]]
[1] 4
```

- Can be used to apply a function to multiple sets of arguments.

```{r}
noise <- function(n, mean, sd) {
      rnorm(n, mean, sd)
}
noise(5, 1, 2)
[1] 0.6355566 1.3298849 2.6023788 0.9256519 1.0427784

noise(1:5, 1:5, 2) #doesn't work properly, doesn't allow vector inputs
[1] 0.9867589 0.4157030 3.0007742 3.7508241 4.7137717

mapply(noise, 1:5, 1:5, 2) #1 var with mean 1, 2 var with mean 2, etc.
[[1]]
[1] -0.2412261

[[2]]
[1]  1.533284 -2.950895

[[3]]
[1] 4.186533 3.268046 3.771591

[[4]]
[1] 6.534970 5.228687 1.867848 4.118644

[[5]]
[1] -0.7377883  4.8755576  3.2978726  3.3713914  0.8926609
```

#### tapply

- Used to apply afunction over subsets of a vector
- Split-Apply-Combine methodology

```{r}
str(tapply)
function (X, INDEX, FUN = NULL, ..., simplify = TRUE)
```

- `X` is a vector
- `INDEX` is a factor or a list of factors (or else they are coerced to factors)
- `FUN` is a function to be applied
- `...` contains other arguments to be passed `FUN`
- `simplify`, should we simplify the result? (like sapply)

```{r}
x <- c(rnorm(10), runif(10), rnorm(10, 1)) 
#10 normals, 10 uniforms, 10 normals with mean = 1
f <- gl(3, 10) #A factor variable

f
 [1] 1 1 1 1 1 1 1 1 1 1 2 2 2 2 2 2 2 2 2 2 3 3 3 3 3 3 3 3 3 3
Levels: 1 2 3

tapply(x, f, mean) #Allows you to take the mean of each group of numbers in 'x'
          1           2           3 
-0.02428736  0.50305875  0.81895348 
          
tapply(x, f, mean, simplify = FALSE) #No simplify gives a list
$`1`
[1] -0.02428736

$`2`
[1] 0.5030588

$`3`
[1] 0.8189535

tapply(x, f, range)
$`1`
[1] -2.281810  2.136192

$`2`
[1] 0.02626713 0.99108003

$`3`
[1] -1.494296  2.228803
```

```{r}
tapply(flags$animate, flags$landmass, mean)
#Splits the animate column into groups based on landmass then calculates the mean for each group.
```

#### split

- Takes a vector or other objects and splits it into groups determined by a factor or list of factors.

```{r}
str(split)
function (x, f, drop = FALSE, ...)
```

- `x` is a vector (or list) or data frame
- `f` is a factor (or coerced to one) or a list of factors
- `drop` indicates whether empty factors levels should be dropped

```{r}
x <- c(rnorm(10), runif(10), rnorm(10, 1))
f <- gl(3, 10)
split(x, f)
$`1`
 [1] -0.63781689 -0.05339353 -1.06600999 -0.14100772  2.46366537
 [6] -0.85588173 -0.63717026  0.88631349 -0.47038351  0.49418802

$`2`
 [1] 0.5994623 0.3907709 0.2096855 0.3136003 0.9593178 0.2082625
 [7] 0.3876043 0.8196400 0.4331901 0.4623739

$`3`
 [1]  1.3042264  0.8355857 -0.6364617  0.7400829  1.7588725  1.0186106
 [7]  1.9998755  1.0091212  1.5489539  1.4735341
```

```{r}
lapply(split(x, f), mean)
$`1`
[1] -0.001749676

$`2`
[1] 0.4783908

$`3`
[1] 1.10524
```

- It's common to use `split` with the `lapply` function
- In this case the `tapply` function can do the exact same thing.
- Always returns a list back (This list contains 3 elements)

```{r}
library(datasets)
head(airquality)
  Ozone Solar.R Wind Temp Month Day
1    41     190  7.4   67     5   1
2    36     118  8.0   72     5   2
3    12     149 12.6   74     5   3
4    18     313 11.5   62     5   4
5    NA      NA 14.3   56     5   5
6    28      NA 14.9   66     5   6
```

```{r}
s <- split(airquality, airquality$Month)
lapply(s, function(x) colMeans(x[, c("Ozone", "Solar.R", "Wind")]))
$`5`
   Ozone  Solar.R     Wind 
      NA       NA 11.62258 

$`6`
    Ozone   Solar.R      Wind 
       NA 190.16667  10.26667 

$`7`
     Ozone    Solar.R       Wind 
        NA 216.483871   8.941935 

$`8`
   Ozone  Solar.R     Wind 
      NA       NA 8.793548 

$`9`
   Ozone  Solar.R     Wind 
      NA 167.4333  10.1800 
```

- Here the factor used to split the data is the "Month" variable and returns a list

```{r}
sapply(s, function(x) colMeans(x[, c("Ozone", "Solar.R", "Wind")]))
               5         6          7        8        9
Ozone         NA        NA         NA       NA       NA
Solar.R       NA 190.16667 216.483871       NA 167.4333
Wind    11.62258  10.26667   8.941935 8.793548  10.1800

sapply(s, function(x) colMeans(x[, c("Ozone", "Solar.R", "Wind")], na.rm = TRUE))
                5         6          7          8         9
Ozone    23.61538  29.44444  59.115385  59.961538  31.44828
Solar.R 181.29630 190.16667 216.483871 171.857143 167.43333
Wind     11.62258  10.26667   8.941935   8.793548  10.18000
```

- `sapply` simplifies the result into a matrix since each element of the list is the same length.

#### Splitting on More than One Level

```{r}
x <- rnorm(10)
f1 <- gl(2, 5)
f2 <- gl(5, 2)

f1
 [1] 1 1 1 1 1 2 2 2 2 2
Levels: 1 2

f2
 [1] 1 1 2 2 3 3 4 4 5 5
Levels: 1 2 3 4 5

interaction(f1, f2)
 [1] 1.1 1.1 1.2 1.2 1.3 2.3 2.4 2.4 2.5 2.5
Levels: 1.1 2.1 1.2 2.2 1.3 2.3 1.4 2.4 1.5 2.5
```

- The interaction function combines all the levels of the first factor with all the levels of the second factor. 2 x 5 = 10

```{r}
str(split(x, list(f1, f2)))
List of 10
 $ 1.1: num [1:2] 0.911 -1.551
 $ 2.1: num(0) 
 $ 1.2: num [1:2] 1.193 0.379
 $ 2.2: num(0) 
 $ 1.3: num 1.2
 $ 2.3: num -0.162
 $ 1.4: num(0) 
 $ 2.4: num [1:2] 0.513 -2.542
 $ 1.5: num(0) 
 $ 2.5: num [1:2] -0.754 0.782

str(split(x, list(f1, f2), drop = TRUE))
List of 6
 $ 1.1: num [1:2] 0.911 -1.551
 $ 1.2: num [1:2] 1.193 0.379
 $ 1.3: num 1.2
 $ 2.3: num -0.162
 $ 2.4: num [1:2] 0.513 -2.542
 $ 2.5: num [1:2] -0.754 0.782
```

- the `drop` argument drops the factors that result in empty levels.

### Debugging
  
Indications that somemthing's not right:
  
- `message:` A generic notification/diagnostic message produced by the `message` function 
  - Execution of the function continues
- `warning:` An indication that something is wrong but not necessarily fatal 
  - Execution of the function continues; generated by the `warning` function
- `error:` An indication that a fatal problem has occured
  - Execution stops
  - Produced by the `stop` function
- `condition:` A generic concept for indicating that something unexpected can occur
  - Programmers can create their own conditions
  
```{r}
log(-1)
[1] NaN
Warning message:
  In log(-1) : NaNs produced
```

```{r}
printmessage <- function(x) {
        if(x > 0)
                print("x is greather than zero")
        else
                print("x is less than or equal to zero")
        invisible(x) #prevents auto-printing
}

printmessage(1)
[1] "x is greather than zero"

#Fix this error

printmessage(NA)
Error in if (x > 0) { : missing value where TRUE/FALSE needed}

printmessage2 <- function(x) {
        if(is.na(x))
                print("x is a missing value!")
        else if(x > 0)
                print("x is greather than zero")
        else
                print("x is less than or equal to zero")
        invisible(x) #prevents auto-printing
}

x <- log(-1)
Warning message:
  In log(-1) : NaNs produced

printmessage2(x)
[1] "x is a missing value!"
```

Questions to ask:
  
- What was your input? How di dyou call the function?
- What were you expecting? Output, messages, other results?
- What did you get?
- How does what you get differ from what you were expecting?
- Were your expectations correct in the first place?
- Can you reproduce the problem (exactly)?
  - In random number generation always set the "seed"
  - Networking programming or parallel functions may be unreproducible
  - Problems with getting data over the internet may be hard to reproduce
  
#### Debugging Tools

- `traceback:` prints out the function call stack after an error occurs
  - does nothing if there's no error
- `debug:` flags a function for "debug" mode which allows you to step through exectution of a function one line at a time
- `browser:` suspends the execution of a function wherever it is called and puts the function in debut mode
- `trace:` allows you to insert debugging code into a function at specific places
- `recover:` allows you to modify the error behaviour so that you can browse the function call stack
  
These are interactive tools to help you pick through a function.
  
Alternatively you can insert print/cat statements in the function.

#### traceback

```{r}
mean(x)
Error in mean(x) : object 'x' not found

traceback()
1: mean(x)
```
  
- Must call `traceback` immediately after the error occurs. Only gives you the most recent error.

```{r}
lm(y - x)
Error in eval(expr, envir, enclos) : object 'y' not found

traceback()
4: stats::model.frame(formula = y - x, drop.unused.levels = TRUE)
3: eval(mf, parent.frame())
2: eval(mf, parent.frame())
1: lm(y - x)
```

#### debug

```{r}
debug(lm)
lm(y - x)

#Here it will print out the whole function body code
#Then you'll be in the browser

Browse[2]>

#Just like your R workspace (a workspace embedded in a workspace)
#The environment within this workspace is your function environment

Browse[2]> n

#This goes to the next line.
#Keep doing this until you get to the line where the error occurs
#You can also debug functions within the function by calling them

undebug(lm)
```

#### recover

```{r}
options(error = recover)

read.csv("nosuchfile")
Error in file(file, "rt") : cannot open the connection
In addition: Warning message:
In file(file, "rt") :
  cannot open file 'nosuchfile': No such file or directory

Enter a frame number, or 0 to exit   

1: read.csv("nosuchfile")
2: read.table(file = file, header = header, sep = sep, quote = quote,
3: file(file, "rt")

Selection:
```

#### Summary

- There are three main indications of a problem/condition:
  - `message`, `warning`, `error`
  - only an `error` is fatal
- When analyzing a function with a problem, make sure you can reproduce the problem, clearly state your expectations and how the output differs from your expectation
- Interactive debugging tools `traceback`, `debug`, `browser`, and `recover` can be used to find problematic code in functions
- Debugging tools are not a substitute for thinking!