---
title: "Week 2 - Programming with R"
output:
  html_document:
    df_print: paged
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```
1. Everything that exists is an Object
2. Everything that happens is a function call  
  
Control Structures Control the flow of execution of the program (depending on runtime conditions)

Common Structures:

- if, else: test a condition
- for: execute a loop a fixed # of times
- while: execute a loop while a condition is true
- repeat: execute an infinite loop
- break: break the execution of a loop
- next: skip an iteration of a loop
- return: exit a function

### if

```{r}
if(condition) {
        #do something
} else if(condition2) {
        #do something different (can do any # of these)
} else {
        #do something else
}
```{r}    
if(x > 3) {
        y <- 10
} else {
        y <- 0
}
```
```{r}
y <- if(x > 3) {
        10
} else {
        0
}
```

### for

```{r}
for(i in 1:4) {
        print(x[i])
}
```
````{r}
for(i in seq_along(x)) {
        print(x[i])
}
```
```{r}
for(letter in x) {
        print(letter)
}
```
```{r}
for(i in 1:4) print(x[i])
```
```{r}
for(i in seq_len(nrow(x))) {
        for(j in seq_len(ncol(x))) {
        print(x[i, j])
        }
}
```

- Going past 3 or 4 loops gets confusing, might want to use functions instead

### While
- Be careful, these loops can be infinite if not careful

```{r}
count <- 0
while(count < 10) {
        print(count)
        count <- count + 1
}
```
```{r}
z <- 5
while(z >= 3 && z <= 10) {
        print(z)
        coin <- rbinom(1, 1, 0.5)

        if(coin == 1) { #random walk
                z <- z + 1
        } else {
                z <- z - 1
        }
}
```

### Repeat
- Have to call break at some point
- Probably better to use a for loop

```{r}
Ex. Gradient Descent:
x0 <- 1
tol <- 1e-8 #tolerance

repeat {
        x1 <- computeEstimateofx()

        if(abs(x1 - x0) <tol) {
                breaks
        } else {
                x0 <- x1
        }
}
```

### Next
- Used to skip an iteration of a loop

```{r}
if(i in 1:100) {
        if(i <= 20) {
                #skip the first 20 iterations
                next
        }
        #do something here
}
```

### Return
- similar but exits a function and usually returns something

### Summary
- if, while and for control the flow
- infinite loops should be avoided
- mostly use these in programs, use *apply functions instead in command-line

### Functions

Exs.
```{r}
add2 <- function(x, y) {
        x + y
}
```
```{r}
above10 <- function(x) {
        use <- x > 10 #creates a logical vector
        x[use]
}
```
```{r}
above <- function(x, n = 10) { #default n is 10
        use <- x > n
        x[use]
}
```
```{r}
columnmean <- function(y, removeNA = TRUE) {
        nc <- ncol(y)
        means <- numeric(nc) #empty vector
        for(i in 1:nc) {
                means[i] <- mean(y[, i], na.rm = removeNA)
        }
        means
}
```

- Functions are treated like first class objects
  - You can pass functions as arguments to other functions
  - Functions can be nested (functions inside of functions)
- The return value is the last expression to be evaluated  

#### Arguments

```{r}
f <- function(<arguments>) {
        #Do something interesting
}
```

- Formal arguments are included in the function def.
- `formals` function returns a list of all the formal arguments of a function.
- Not every function call in R makes use of all the formal arguments.
- Function arguments can be missing or might have default values
- Arguments can be matched positionally or by name.
  - Different positions is not recommended as it can lead to confusion.
  - When an argument is matched by name, it is *taken out* of the argument list.
  - The remaning arguments are matched in the function order.
  - Partial name matching also works.
  
```{r}
args(lm)
function (formula, data, subset, weights, na.action,
          method = "qr", model = TRUE, x = FALSE, 
          y = FALSE, qr = TRUE, singular.ok = TRUE,
          contrasts = NULL, offset, ...)
```
The following two calls are equivalent:
```{r}
lm(data = mydata, y - x, model = FALSE, 1:100)
lm(y - x, mydata, 1:100, model = FALSE)
```

#### Defining a Function

##### Lazy Evaluation  

```{r}
f <- function(a, b) {
    a^2
}
f(2)
```
- Argument `b` is useless, so calling `f(2)` will not produce an error  
```{r}
f <- function(a, b) {
    print(a)
    print(b)
}
f(45)

## [1] 45

## Error: argument "b" is missing, with no default
```
- Error occurs after printing `a`.  
  
##### ... Argument

- Used to indicate a variable number of the arguments.
- Some generic functions use this so that extra arguments to be passed in.
- Useful when extending a function.

```{r}
myplot <- function(x, y, type = "l", ...) {
        plot(x, y, type = type, ...)
}
```

- Sometimes a function takes an unknown number of arguments as input.

```{r}
args(paste)
function (..., sep = " ", collapse = NULL)
  
args(cat)
function (..., file = "", sep = " ", fill = FALSE, labels = NULL, append = FALSE)
```

- Any arguments that appear after ... must be named explicitely and cannot be partially matched.

```{r}
args(paste)
function (..., sep = " ", collapse = NULL)
  
paste("a", "b", sep = ":")
[1] "a:b"

paste("a", "b", se = ":")
[1] "a b :"
```

### Binding Values to Symbols

- Conflicting definitions:

```{r}
lm <- function(x) { x * x }
```

- How does R know what value to assign lm (It is already defined in the *stats* package)?
- R searches through a series of `environments` to find the appropriate value.
1. Search the global environment
2. Search the namespaces of each package on the search list
- The search priority can be found using the `search` function, and this priority changes based on when you load the packages.
- The global environment (user's workspace) is always the first element of the search list and the base package is always last.
- When a user loads a package with `library` the namespace of that package gets put in position 2.
- Note that R has separate namespaces for functions and non-functions (ie. you can have a function `c` and a variable `c`).

### Scoping Rules

- Main feature that separates R from S
- They determine how a value is associated with a free variable in a function
- R uses *lexical Scoping* or *static scoping*. A common alternative is *dynamic scoping*
- Related to the scoping rules is how R uses the search *list* to bind a value to a symbol.
- Lexical scoping turns out to be particularly useful for simplifying statistical computations

#### Lexical Scoping

```{r}
f <- function(x, y) {
        x^2 + y / z
}
```
- Here `z` is called a *free variable*
- Scoping rules determine how values are assigned to free variables
- Free variables are not formal arguments and ar not local variables (from within the function body).

*The values of free variables are searched for in the environment in which the function was defined.*
  
What is an environment?
  
- A collection of (symbol, value) pairs, i.e. x is a symbol and 3.14 might be its value. (Everything in R is a symbol and a value)
- Every environment has a parent environment; it is possible for an environment to have multiple "children"
- The only environment without a parent is the empty environment
- A function + an environment = a closure or function closure.
  - Key to a lot of interesting operations in R
  
Searching for the value of a free variable:
  
- If the value of a symbol is not found in the environment in which a function was defined, then the search is continued in the parent environment.
- The search continues down the sequence of parent environments until we hit the top-level environment; this is usually the global environment (workspace) or the namespace of a package.
- After the top-level environment, the search continues down the search list until we hit the empty environment. If a value for a given symbol cannot be found once the empty environment is arrived at, then an error is thrown.
  
Why does this even matter?
  
- Typically, a function is defined in the global environment, so that the values of free variables are just found in the user's workspace. 
  - This behavior is logical for most people and is usually the "right thing" to do
- However, in R you can have functions defined *inside other functions* 
  - A function can return another function
  - Languages like C don't let you do this
- Now things get interesting -- In this case the environment in which a function is defined is the body of another function!

```{r}
make.power <- function(n) {
        pow <- function(x) {
                x^n
        }
        pow
}
```
- This function returns another function as its value.
```{r}
cube <- make.power(3) #Returns pow <- function(x) {x^3}
square <- make.power(2) #Returns pow <- function(x) {x^2}
#These lines define the variable n and become a new function that can define x

cube(3)
[1] 27

square(3)
[1] 9
```

```{r}
ls(environment(cube))
[1] "n" "pow"
get("n", environment(cube))
[1] 3

ls(environment(square))
[1] "n" "pow"
get("n", environment(square))
[1] 2
```

```{r}
y <- 10
f <- function(x) {
        y <- 2
        y^2 + g(x)
}

g <- function(x) {
        x*y
}

f(3)
#What is this output supposed to be?

```
  
#### Lexical vs. Dynamic Scoping

Lexical Scoping:  

- The value of `y` in the function `g` is looked up in the environment in which the function was defined.
  - In this case, the global environment. 
  - `y = 10`
  
Dynamic Scoping:  

- The value of `y` is looked up in the environment from which the function was called 
  - Referred to as the calling environment.
  - In R the calling environment is known as the parent frame
  - `y = 2`
  
```{r}
g <- function(x) {
  a <- 3
  x+a+y
}

g(2)
Error in g(2) : object "y" not found

y <- 3
g(2)
[1] 8
```
  
- The function is defined in the global environment and is subsequently called from the global environment.
- The defining environment and the calling environment are the same.
- Sometimes gives the appearance of dynamic scoping.
  
#### Other Languages
- Scheme
- Perl
- Python
- Common Lisp
  
#### Consequences of Lexical Scoping

- In R all objects must be stored in memory.
  - A challenge with large datasets.
- All functions must carry a pointer to their respective defining environments, which could be anywhere.
- In S-PLUS, free variables are always looked up in the global workspace, so everything can be stored on the disk because the "defining environment" of all functions is the same.  

#### Application: Optimization

- Optimization routines in R like `optim`, `nlm`, and `optimize` require you to pass a function whose argument is a vector of parameters (e.g. a log-likelihood)
- However, an object function might depend on a host of other things besides its parameters (like data)
- When writing software which does optimization, it may be desirable to allow the user to hold certain parameters fixed.
  
Constructor function:
```{r}
make. NegLogLik <- function(data, fixed=c(FALSE,FALSE)) {
        params <- fixed
        function(p) {
                params[!fixed] <- p
                mu <- params[1]
                sigma <- params[2]
                a <- -0.5*lengthy(dat)*log(2*pi*sigma^2)
                b <- 0.5*sum((data-mu)^2) / (sigma^2)
                -(a + b)
        }
}
```
  
Maximizingn a Normal Likelihood:
```{r}
set.seed(1); normals <- rnorm(100, 1, 2)
nLL <- make.NegLogLik(normals)
nLL
function(p) {
        params[!fixed] <- p
        mu <- params[1]
        sigma <- params[2]
        a <- -0.5*length(data)*log(2*pi*sigma^2)
        b <- -0.5*sum((data-mu)^2) / (sigma^2)
        -(a + b)
}
<environment: 0x165b1a4> 
#Enclosing environment for this function.
# This would not be present if you were defining in the global environment.  
ls(environment(nLL))
[1] "data"  "fixed" "params"
```
  
Estimating Parameters:
```{r}
optim(c(mu = 0, sigma = 1), nLL)$par
      mu    sigma
  1.218239  1.787343
```
  
Fixing sigma = 2
```{r}
nLL <- make.NegLogLik(normals, c(FALSE, 2))
optimize(nLL, c(-1, 3))$minimum
[1] 1.800596
```
  
Fixing mu = 1
```{r}
nLL <- make.NegLogLik(normals, c(1, FALSE))
optimize(nLL, c(1e-6, 10))$minimum
[1] 1.800596
```
  
Plotting the Likelihood:
```{r}
nLL <- make.NegLogLik(normals, c(1, FALSE))
x <- seq(1.7, 1.9, len = 100)
y <- sapply(x, nLL)
plot(x, exp(-(y - min(y))), type = "l")

nLL <- make.NegLogLik(normals, c(FALSE, 2))
x <- seq (0.5, 1.5, len = 100)
y <- sapply (x, nLL)
plot(x, exp(-(y - min(y))), type = "l")
```
  
#### Summary
- Objective functions can be "built" which contain all of the necessary data for evaluating the function
- No need to carry around long argument lists -- useful for interactive and exploratory work.
- Code can be simplified and cleaned up
  
### Coding Standards

1. Always write in a text editor and make a text file
2. Indent your code (4 - 8 spaces)
3. Limit the width of your code to ~80 columns
4. Limit length of individual functions
  - Have separate functions for reading data, analyzing, plotting, etc.
  - When a bug occurs it'll be easier to find it to fix it.
  
### Dates and Times

- Dates are represented by the `Date` class
  - Stored internally as the number of days since 1970-01-01
- Times are represented by the `POSIXct` or the `POSIXlt` class
  - Stored internally as the number of seconds since 1970-01-01

```{r}
x <- as.Date("1970-01-01")
x
## [1] "1970-01-01"
unclass(x)
## [1] 0
## 0 days since 1970-01-01
unclass(as.Date("1970-01-02"))
## [1] 1
```
  
- `POSIXct` is just a very large integer under the hood; it uses a useful class when you want to store times in something like a data frame.
- `POSIXlt` is a list underneath and it stores a bunch of other useful information like the day of the week, day of the year, month, day of the month  

Generic Functions:  

- `weekdays`: give the day of the week
- `months`: give the month name
- `quarters`: give the quarter number ("Q1", "Q2", etc.) 

```{r}
x <- Sys.time()
x #In POSIXct format
## [1] "2013=01=24 22:04:14 EST"
unclass(x)
## [1] 1349083054
x$sec
## Error: $ operator is invalid for atomic vectors

p <- as.POSIXlt(x)
names(unclass(p))
## [1] "sec"  "min"  "hour"  "mday"  "mon"
## [6] "year"  "wday"  "yday"  "isdst"
p$sec
## [1] 14.34
```

```{r}
datestring <- c("January 10, 2012 10:40", 'December 9, 2011 9:10")
x <- strptime(datestring, "%B %d, %Y %H:%M")
x
[1] "2012-01-10 10:40:00 EST" "2011-12-09 09:10:00 EST"

class(x)
[1] "POSIXlt" "POSIXt"
```
  
Operations on Dates and Times:
```{r}
x <- as.Date9"2012-01-01")
y <- strptime("9 Jan 2011 11:34:21", "%d %b %Y %H:%M:%S")
x-y
## Warning: Incompatible methods ("-.Date", "-.POSIXt") for "-"
## Error: non-numeric argument to binary operator
x <- as.POSIXlt(x)
x-y
## Time difference of 356.3 days
```
- Can also take care of leap years and time zones (tz)  

